Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-11-22 Manuscript/Manuscript.Rmd
Update md pattern, add new case study

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(
    pat = rep("systematic", 5),
    rownr = 5:1,
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(X1, X2, X3)),
  tibble(
    pat = rep("sporadic", 5),
    rownr = 5:1,
    X1 = c("observed", "missing", rep("observed", 3)),
    X2 = c(rep("observed", 2), rep("missing", 3)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(X1, X2, X3))
) %>%
  mutate(across(everything(), as.factor),
         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE))
# plot
ggplot(patterns, aes(
  x = name,
  y = rownr,
  fill = value,
  type = pat
)) +
  geom_tile(color = "black") +
  scale_y_discrete(labels = c(2, 2, 1, 1, 1)) +
  scale_fill_manual(values = plot_col, name = "Datapoint") +
  facet_wrap( ~ pat, ncol = 2) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "Variable", y = "Cluster")
```

@@

We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).

@@

```{r impact, echo=FALSE}
# load data
data("impact")
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# missingness
plot_md_pat(impact)
```
-> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf), the data is already imputed (Steyerberg et al, 2008).

@@

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# load data
pop <- readRDS("../Data/popNCR.RDS")
# ICCs
icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
```
- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`.

@@

```{r pop_pat, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
```

- Does the missing data of popular depend on popteach? One could for example check this by making a histogram of popteach separately for the pupils with known popularity and missing popularity. We do see that the histogram for the missing popular (TRUE) is further to the right than the histogram for observed popular (FALSE). This would indicate a right-tailed MAR missingness. In fact this is exactly what happens, because we created the missingness in these data ourselves. But we can make it observable by examining the relations between the missingness in popular and the observed data in popteach.

- Does the missingness in teacher popularity depend on pupil popularity? Yes: there is a dependency. The relation seems to be right-tailed.

```{r pop_dist, echo=FALSE, message=FALSE, warning=FALSE}
# distributions
plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
```

@@

- Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well. Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeniality, but don't mention the term) ->  Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there. Sometimes there's little reason to assume some variable is affected by heterogeneity. -> Refer to Meng, Vincent, and a paper by Grund on congeniality and random slopes.

- Step 0: As predictor

@@

## Step 1-3

>>>>

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(
    pat = rep("systematic", 5),
    rownr = 5:1,
    # obs = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3)),
  tibble(
    pat = rep("sporadic", 5),
    rownr = 5:1,
    # obs = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = c("observed", "missing", rep("observed", 3)),
    X2 = c(rep("observed", 2), rep("missing", 3)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3))
) %>%
  mutate(across(everything(), as.factor),
         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE),
         text = rep(c(rep(c("1","", "", ""), 3), rep(c("2","", "", ""), 2)),2))
# plot
ggplot(patterns, aes(
  x = name,
  y = rownr,
  color = value,
  type = pat,
  width = 0.8, 
  height = 0.8,
  label = text
)) +
  geom_tile(fill = "white", size = 2) +
  geom_text() +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = 5:1) +
  scale_color_manual(values = plot_col, name = "Datapoint") +
  facet_wrap( ~ pat, ncol = 2, strip.position = "bottom") +
  theme_minimal() +
  theme(
    legend.position = "none", 
    panel.grid.major = element_blank()) +
  labs(x = "", y = "Observation (e.g., patient)")
# TODO: add patient nr and cluster
# TODO: add grid.lines = "none"

```

@@

We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)). -> ask whether we can use the Heckman repo data or simulate data ourselves

Heckman options:

- leiden85

- GJRM::hiv (https://rdrr.io/github/egeminiani/GJRM/man/hiv.html)

- simulating

- IMPACT

@@

```{r impact, echo=FALSE}
# load data
data("impact")
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# missingness
plot_md_pat(impact)
# TODO: rotate variable labels
```
-> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf), the data is already imputed (Steyerberg et al, 2008).

- MAR miss varying by cluster. Obs data patt differ per cluster. E.g., in cluster 1 miss depends on age but not in cluster two. Split the dataframe and run `ampute()` on each cluster. -> TODO: also make MNAR missingness to heckman model. Maybe based on `ct` variable? Inclusion-selection variable. -> otherwise: use `leiden85` data on blood pressure with MNAR. Then run cox regression like the boshuizen article but with living situation as clusters. -> TODO: get analyses from https://www.gerkovink.com/mimp/Contents/Exercises/Day%203%20-%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html.

- ADD: `multilevel_ampute()` wrapper function in `mice`.

@@

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# load data
pop <- readRDS("../Data/popNCR.RDS")
# ICCs
icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
```
- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would be wise to use multilevel modeling.

@@

```{r pop_pat, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

- Does the missing data of `popular` depend on `popteach`? One could for example check this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity. And the other way around: Does the missingness in teacher popularity depend on pupil popularity?

```{r pop_dist, echo=FALSE, message=FALSE, warning=FALSE}
# distributions
plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
# TODO: check if up-side-down plot works
# or maybe overlapping densities
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
```

-  We do see that the histogram for the missing `popular` is further to the right than the histogram for observed `popular`. This would indicate a right-tailed MAR missingness. In fact this is exactly what happens, because we created the missingness in these data ourselves. But we made it observable by examining the relations between the missingness in popular and the observed data in `popteach`. There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed.

## Amputation

-

@@

- Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well. Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeniality, but don't mention the term) ->  Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there. Sometimes there's little reason to assume some variable is affected by heterogeneity. -> Refer to @meng94, an Audigier paper, and a paper by Grund on congeniality and random slopes.

- Step 0: As predictor + CCA to scare off users

@@

## Step 1-3 + MNAR

### 2021-11-22 Manuscript/Manuscript.tex
Update md pattern, add new case study

We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
version of the \texttt{mice::popmis} data (i.e., a variation on the Hox
(2010) popularity data, where the missingness in the variables is either
missing at random (MAR) or missing not at random (MNAR)).

@@

@@

  \begin{itemize}
  \tightlist
  \item
    \texttt{pupil} Pupil number within class,
  \item
    \texttt{class} Class number,
  \item
    \texttt{extrav} Pupil extraversion,
  \item
    \texttt{sex} Pupil gender,
  \item
    \texttt{texp} Teacher experience (years),
  \item
    \texttt{popular} Pupil popularity,
  \item
    \texttt{popteach} Teacher popularity.
  \end{itemize}
\item
  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
  \texttt{popteach} it is 0.31.
\item
  What does the missingness look like? Induced MAR/MNAR missingness.
  Missing data pattern:
\end{itemize}

@@

\begin{itemize}
\item
  Does the missing data of popular depend on popteach? One could for
  example check this by making a histogram of popteach separately for
  the pupils with known popularity and missing popularity. We do see
  that the histogram for the missing popular (TRUE) is further to the
  right than the histogram for observed popular (FALSE). This would
  indicate a right-tailed MAR missingness. In fact this is exactly what
  happens, because we created the missingness in these data ourselves.
  But we can make it observable by examining the relations between the
  missingness in popular and the observed data in popteach.
\item
  Does the missingness in teacher popularity depend on pupil popularity?
  Yes: there is a dependency. The relation seems to be right-tailed.
\end{itemize}

@@

@@

\begin{itemize}
\item
  Which models will we discuss? We'll build the model to grow in
  complexity. The final model is the most complex but also the most
  versatile.
\item
  Note on model complexity: Typically, we should at least use random
  intercepts, but often random slopes as well. Ideally we impute with
  random everything and heteroscedastic errors: most generic method (no
  worry about congeniality, but don't mention the term) -\textgreater{}
  Refer to other papers for background, we'll focus just on the software
  implementation of the situations mentioned there. Sometimes there's
  little reason to assume some variable is affected by heterogeneity.
  -\textgreater{} Refer to Meng, Vincent, and a paper by Grund on
  congeniality and random slopes.
\item
  Step 0: As predictor
\item
  Step 1: Random intercepts
\item
  Step 2: Random slopes
\item
  Step 3: Residuals
\item
  Heckman model for MNAR
\item
  What do the different implementations look like? How to define the
  imputation model(s) in \texttt{mice}?
\end{itemize}

@@

\hypertarget{step-1-3}{%
\subsection{Step 1-3}\label{step-1-3}}

>>>>

We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
version of the \texttt{mice::popmis} data (i.e., a variation on the Hox
(2010) popularity data, where the missingness in the variables is either
missing at random (MAR) or missing not at random (MNAR)).
-\textgreater{} ask whether we can use the Heckman repo data or simulate
data ourselves

Heckman options:

\begin{itemize}
\item
  leiden85
\item
  GJRM::hiv
\item
  simulating
\item
  IMPACT
\end{itemize}

@@

\begin{itemize}
\item
  MAR miss varying by cluster. Obs data patt differ per cluster. E.g.,
  in cluster 1 miss depends on age but not in cluster two. Split the
  dataframe and run \texttt{ampute()} on each cluster. -\textgreater{}
  TODO: also make MNAR missingness to heckman model. Maybe based on
  \texttt{ct} variable? Inclusion-selection variable. -\textgreater{}
  otherwise: use \texttt{leiden85} data on blood pressure with MNAR.
  Then run cox regression like the boshuizen article but with living
  situation as clusters. -\textgreater{} TODO: get analyses from
  \url{https://www.gerkovink.com/mimp/Contents/Exercises/Day\%203\%20-\%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html}.
\item
  ADD: \texttt{multilevel\_ampute()} wrapper function in \texttt{mice}.
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    \texttt{pupil} Pupil number within class,
  \item
    \texttt{class} Class number,
  \item
    \texttt{extrav} Pupil extraversion,
  \item
    \texttt{sex} Pupil gender,
  \item
    \texttt{texp} Teacher experience (years),
  \item
    \texttt{popular} Pupil popularity,
  \item
    \texttt{popteach} Teacher popularity.
  \end{itemize}
\item
  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
  \texttt{popteach} it is 0.31. It would be wise to use multilevel
  modeling.
\item
  What does the missingness look like? Induced MAR/MNAR missingness.
  Missing data pattern:
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Does the missing data of \texttt{popular} depend on \texttt{popteach}?
  One could for example check this by making a histogram of
  \texttt{popteach} separately for the pupils with known popularity and
  missing popularity. And the other way around: Does the missingness in
  teacher popularity depend on pupil popularity?
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  We do see that the histogram for the missing \texttt{popular} is
  further to the right than the histogram for observed \texttt{popular}.
  This would indicate a right-tailed MAR missingness. In fact this is
  exactly what happens, because we created the missingness in these data
  ourselves. But we made it observable by examining the relations
  between the missingness in popular and the observed data in
  \texttt{popteach}. There is also a dependency between the missingness
  in teacher popularity and pupil popularity. The relation seems to be
  right-tailed.
\end{itemize}

\hypertarget{amputation}{%
\subsection{Amputation}\label{amputation}}

\begin{itemize}
\tightlist
\item
\end{itemize}

@@

\begin{itemize}
\item
  Which models will we discuss? We'll build the model to grow in
  complexity. The final model is the most complex but also the most
  versatile.
\item
  Note on model complexity: Typically, we should at least use random
  intercepts, but often random slopes as well. Ideally we impute with
  random everything and heteroscedastic errors: most generic method (no
  worry about congeniality, but don't mention the term) -\textgreater{}
  Refer to other papers for background, we'll focus just on the software
  implementation of the situations mentioned there. Sometimes there's
  little reason to assume some variable is affected by heterogeneity.
  -\textgreater{} Refer to \citet{meng94}, an Audigier paper, and a
  paper by Grund on congeniality and random slopes.
\item
  Step 0: As predictor + CCA to scare off users
\item
  Step 1: Random intercepts
\item
  Step 2: Random slopes
\item
  Step 3: Residuals
\item
  Heckman model for MNAR
\item
  What do the different implementations look like? How to define the
  imputation model(s) in \texttt{mice}?
\end{itemize}

@@

\hypertarget{step-1-3-mnar}{%
\subsection{Step 1-3 + MNAR}\label{step-1-3-mnar}}
